---
title: "Parametric Bootstrap of an MGPM"
author: "Venelin Mitov"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Performing Parametric Bootstrap of an MGPM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
library(data.table)
library(PCMBase)
library(PCMFit)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if(!requireNamespace("ggtree")) {
  message("Building the vignette requires ggtree R-package. Trying to install.")
  status.ggtree <- try({
    if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager")
    BiocManager::install("ggtree", version = "3.8")
  }, silent = TRUE)
  if(class(status.ggtree == "try-error")) {
    stop(
      "The ggtree installation did not succeed. The vignette cannot be built.")
  }
}

generateParametricBootstrapDatasets <- FALSE
```

**Note:** *The writing of this vignette is currently in progress. Please, contact the author for assistance, in case you need to use PCMFit immediately and cannot find your way through the coding examples. Thanks for your understanding.*

# Simulating parametric bootstrap datasets

```{r, eval=generateParametricBootstrapDatasets}
# make results reproducible
set.seed(12, kind = "Mersenne-Twister", normal.kind = "Inversion")

numBootstraps <- 10

bestModel <- 
  RetrieveBestFitScore(PCMFitDemoObjects$fitMGPM_A_F_BC2_RR)$inferredModel

metaI <- PCMInfo(
  X = NULL, tree = attr(bestModel, "tree"), model = bestModel)

# Simulate data with the best fit model from the mammal data with SEs
valuesBootstrapMGPM_A_F_BC2_RR <- data.table(
  IdGlob = seq_len(numBootstraps),
  X = lapply(seq_len(numBootstraps), function(id) {
    PCMSim(
      tree = attr(bestModel, "tree"), 
      model = bestModel, 
      X0 = bestModel$X0,
      metaI = metaI)
  })
)

options(PCMBase.Value.NA = -1e20)
options(PCMBase.Lmr.mode = 11)
options(PCMBase.Threshold.EV = 1e-7)

valuesBootstrapMGPM_A_F_BC2_RR[
  , c("df", "logLik", "score") := {
    resLists <- lapply(.I, function(i) {
      attr(bestModel, "X") <- X[[i]]
      ll <- logLik(bestModel)
      aic <- AIC(bestModel)
      df <- attr(ll, "df")
      c(df, ll, aic)
    })
    list(df = sapply(resLists, function(.) .[1]),
         logLik = sapply(resLists, function(.) .[2]),
         score = sapply(resLists, function(.) .[3]))
  }]

PCMFitDemoObjects$valuesBootstrapMGPM_A_F_BC2_RR <- valuesBootstrapMGPM_A_F_BC2_RR
usethis::use_data(PCMFitDemoObjects, overwrite = TRUE)
```

# Model inference on the parametric bootstrap datasets

## Writing a parametric bootstrap inference R-script
```{r, results="hide", warning=FALSE, message=FALSE, eval=FALSE}
# File: FitBootstrap_MGPM_A_F_BC2_RR.R
# Usage: R --vanilla --slave -f ../../FitBootstrap_MGPM_A_F_BC2_RR.R --args 1
library(PCMBase)
library(PCMBaseCpp)
library(PCMFit)
library(data.table)

# extract dataset identifier and possibly other parameters from the command line:
args <- commandArgs(trailingOnly = TRUE)
if(length(args) > 0) {
  data_id <- as.integer(args[1])
} else {
  data_id <- 1L
}

# A character string used in filenames for a model inference on a given data:
prefixFiles = paste0("MGPM_A_F_BC2_RR_BSID_", data_id)

# creating the cluster for this PCMFit run:
if(!exists("cluster") || is.null(cluster)) {
  if(require(doMPI)) {
    # using MPI cluster as distributed node cluster (possibly running on a 
    # cluster of multiple nodes)
    # Get the number of cores. Assume this is run in a batch job.
    p = strtoi(Sys.getenv('LSB_DJOB_NUMPROC'))
    cluster <- startMPIcluster(count = p-1, verbose = TRUE)
    doMPI::registerDoMPI(cluster)
  } else {
    # possibly running on personal computer without mpi installation
    cluster <- parallel::makeCluster(
      parallel::detectCores(logical = TRUE),
      outfile = paste0("log_", prefixFiles, ".txt"))
    doParallel::registerDoParallel(cluster)
  }
}

# This function is going to be executed on each worker node.
generatePCMModelsFunction <- function() {
  # make results reproducible
  set.seed(4, kind = "Mersenne-Twister", normal.kind = "Inversion")

  PCMGenerateModelTypes()
  fileName <- '../../DefineParameterLimits.R'
  codeDefineLimits <- readChar(fileName, file.info(fileName)$size)
  eval(parse(text = codeDefineLimits), .GlobalEnv)
}

bestModel <- 
  RetrieveBestFitScore(PCMFitDemoObjects$fitMGPM_A_F_BC2_RR)$inferredModel

tree <- PCMTree(PCMFitDemoObjects$dtSimulated$tree[[1]])
X <- PCMFitDemoObjects$valuesBootstrapMGPM_A_F_BC2_RR$X[[data_id]][
  , seq_len(PCMTreeNumTips(tree))]

currentResultFile <- paste0("CurrentResults_fits_", prefixFiles, ".RData")
if(file.exists(currentResultFile)) {
  load(currentResultFile)
  tableFitsPrev <- listResults$tableFits
} else {
  tableFitsPrev <- NULL
}

fitMGPM_A_F_BC2_RR <- PCMFitMixed(
    X = X, tree = tree, metaIFun = PCMInfoCpp,
    generatePCMModelsFun = generatePCMModelsFunction, 
    maxNumRoundRobins = 2, maxNumPartitionsInRoundRobins = 2,
    tableFitsPrev = tableFitsPrev,
    prefixFiles = prefixFiles,
    doParallel = TRUE)

save(fitMGPM_A_F_BC2_RR, file = paste0("Result_", prefixFiles, ".RData"))
```

## Running the parametric bootstrap inference script on a cluster
```{sh eval=FALSE}
# File: FitBootstrap_MGPM_A_F_BC2_RR_bsub.sh
# Usage: sh FitBootstrap_MGPM_A_F_BC2_RR_bsub.sh
for id in `seq 1 1 10`
do
mkdir -p ResultsBootstrap/MGPM_A_F_BC2_RR_BSID_$id
cd ResultsBootstrap/MGPM_A_F_BC2_RR_BSID_$id
if [ -f "Result_MGPM_A_F_BC2_RR_BSID_"$id".RData" ]
then
rm MPI*.log
rm Current*.RData
else
bsub -M 10000 -n 8 -W 3:59 -R ib sh R --vanilla --slave -f ../../FitBootstrap_MGPM_A_F_BC2_RR.R --args $id
fi
cd ../..
done
```

# Collect bootstrap results
```{r, eval=TRUE, include=FALSE}
# model and tree used to generate the bootstrap datasets
inferredModel <- 
  RetrieveBestFitScore(PCMFitDemoObjects$fitMGPM_A_F_BC2_RR)$inferredModel
inferredTree <- PCMTree(attr(inferredModel, "tree"))
epochsHD <- seq(.2, 
                max(PCMTreeNodeTimes(inferredTree)), 
                by = 2)

if(is.null(PCMFitDemoObjects$bs_MGPM_A_F_BC2_RR)) {
  bs_MGPM_A_F_BC2_RR <- CollectBootstrapResults(
    resultDir = "ResultsBootstrap",
    prefixFiles = "MGPM_A_F_BC2_RR_BSID_",
    nameFitObject = 'fitMGPM_A_F_BC2_RR',
    ids = seq_len(10),
    inferredTree = inferredTree,
    epochs = epochsHD,
    minLength = 0.2,
    verbose = FALSE)
  
  PCMFitDemoObjects$bs_MGPM_A_F_BC2_RR <- bs_MGPM_A_F_BC2_RR
  usethis::use_data(PCMFitDemoObjects, overwrite = TRUE)
} else {
  bs_MGPM_A_F_BC2_RR <- PCMFitDemoObjects$bs_MGPM_A_F_BC2_RR
}
```
  
```{r, eval=FALSE}
inferredModel <- 
  RetrieveBestFitScore(PCMFitDemoObjects$fitMGPM_A_F_BC2_RR)$inferredModel
inferredTree <- PCMTree(attr(inferredModel, "tree"))

epochsHD <- seq(.2, max(PCMTreeNodeTimes(inferredTree)), 
                by = 2)

bs_MGPM_A_F_BC2_RR <- CollectBootstrapResults(
  resultDir = "ResultsBootstrap",
  prefixFiles = "MGPM_A_F_BC2_RR_BSID_",
  nameFitObject = 'fitMGPM_A_F_BC2_RR',
  ids = seq_len(10),
  inferredTree = inferredTree,
  epochs = epochsHD,
  minLength = 0.2,
  verbose = FALSE)
```

# References
