---
title: "Inferring an MGPM with Unknown Shifts"
author: "Venelin Mitov"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Inferring an MGPM with Unknown Shifts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if(!requireNamespace("ggtree")) {
  message("Building the vignette requires ggtree R-package. Trying to install.")
  status.ggtree <- try({
    if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager")
    BiocManager::install("ggtree", version = "3.8")
  }, silent = TRUE)
  if(class(status.ggtree == "try-error")) {
    stop(
      "The ggtree installation did not succeed. The vignette cannot be built.")
  }
}

useCacheMGPM_A_F_BC2_RR <- TRUE
```

**Note:** *The writing of this vignette is currently in progress. If needed, please, contact the author for assistance. Thanks for your understanding.*

# The main R-script
## Loading needed packages and parsing command line arguments
```{r, results="hide", warning=FALSE, message=FALSE, eval=TRUE}
library(PCMBase)
library(PCMBaseCpp)
library(PCMFit)
library(data.table)
# other needed packages, e.g. ape, data.table etc...

# A character string used in filenames for a model inference on a given data:
prefixFiles = paste0("MGPM_A_F_BC2_RR")
```

## Setting up a parallel cluster
### Using MPI
```{r, results="hide", warning=FALSE, message=FALSE, eval=FALSE}
# creating the cluster for this PCMFit run:
if(!exists("cluster") || is.null(cluster)) {
  if(require(doMPI)) {
    # using MPI cluster as distributed node cluster (possibly running on a 
    # cluster of multiple nodes)
    # Get the number of cores. Assume this is run in a batch job.
    p = strtoi(Sys.getenv('LSB_DJOB_NUMPROC'))
    cluster <- startMPIcluster(count = p-1, verbose = TRUE)
    doMPI::registerDoMPI(cluster)
  } else {
    warning("Could not create an MPI cluster")
  }
}
```

### Using a personal computer without mpi installation
```{r, results="hide", warning=FALSE, message=FALSE, eval=FALSE}
if(!exists("cluster") || is.null(cluster)) {
  if(require(parallel)) {
    # possibly running on personal computer without mpi installation
    cluster <- parallel::makeCluster(
      parallel::detectCores(logical = TRUE),
      outfile = paste0("log_", prefixFiles, ".txt"))
    doParallel::registerDoParallel(cluster)
  } else {
    warning("Could not create a parallel cluster")
  }
}
```

## Generating PCMModels on the worker nodes
```{r}
# This function is going to be executed on each worker node.
generatePCMModelsFunction <- function() {
  # make results reproducible
  set.seed(4, kind = "Mersenne-Twister", normal.kind = "Inversion")

  PCMGenerateModelTypes()
  # An example DefineParameterLimits.R file can be found in 
  # vignettes/DefineParameterLimits.R of the PCMFit package source-code. 
  # Note that the path here is relative to the working directory of
  # the worker node R-process. 
  source('DefineParameterLimits.R', local=FALSE)
}
```

## Tree and trait data 
```{r}
tree <- PCMTree(PCMFitDemoObjects$dtSimulated$tree[[1]])
X <- PCMFitDemoObjects$dtSimulated$X[[1]][, seq_len(PCMTreeNumTips(tree))]
```

## Running the MGPM Inference
```{r, include=FALSE, warning=FALSE, results="hide", eval = TRUE}
currentResultFile <- paste0("Current_", prefixFiles, ".RData")
if(file.exists(currentResultFile)) {
  load(currentResultFile)
  tableFitsPrev <- listResults$tableFits
} else {
  tableFitsPrev <- NULL
}

if(is.null(PCMFitDemoObjects$fitMGPM_A_F_BC2_RR) || 
   !useCacheMGPM_A_F_BC2_RR) {
  fitMGPM_A_F_BC2_RR <- PCMFitMixed(
    X = X, tree = tree, metaIFun = PCMInfoCpp,
    tableFitsPrev = tableFitsPrev,
    generatePCMModelsFun = generatePCMModelsFunction, 
    maxNumRoundRobins = 2, maxNumPartitionsInRoundRobins = 2,
    printFitVectorsToConsole = TRUE,
    doParallel = TRUE)
  
  PCMFitDemoObjects$fitMGPM_A_F_BC2_RR <- fitMGPM_A_F_BC2_RR
  usethis::use_data(PCMFitDemoObjects, overwrite = TRUE)
} 

fitMGPM_A_F_BC2_RR <- PCMFitDemoObjects$fitMGPM_A_F_BC2_RR
```

```{r, eval=FALSE}
currentResultFile <- paste0("Current_", prefixFiles, ".RData")
if(file.exists(currentResultFile)) {
  load(currentResultFile)
  tableFitsPrev <- listResults$tableFits
} else {
  tableFitsPrev <- NULL
}

fitMGPM_A_F_BC2_RR <- PCMFitMixed(
    X = X, tree = tree, metaIFun = PCMInfoCpp,
    generatePCMModelsFun = generatePCMModelsFunction, 
    maxNumRoundRobins = 2, maxNumPartitionsInRoundRobins = 2,
    tableFitsPrev = tableFitsPrev,
    prefixFiles = prefixFiles,
    doParallel = TRUE)
```

# Retrieving and analyzing the best scoring model fit
```{r, echo=FALSE}
modelTrue <- PCMFitDemoObjects$dtSimulated$model[[1]]

# We specify the tree and trait values for the true model in order to easily 
# calculate parameter count likelihood and AIC for it:
attr(modelTrue, "tree") <- PCMFitDemoObjects$dtSimulated$treeWithRegimes[[1]]
attr(modelTrue, "X") <- X
attr(modelTrue, "SE") <- X * 0.0
```

```{r}
bestFit <- RetrieveBestFitScore(fitMGPM_A_F_BC2_RR)

modelTrue <- PCMFitDemoObjects$dtSimulated$model[[1]]
# We specify the tree and trait values for the true model in order to easily 
# calculate parameter count likelihood and AIC for it:
attr(modelTrue, "tree") <- PCMFitDemoObjects$dtSimulated$treeWithRegimes[[1]]
attr(modelTrue, "X") <- X
attr(modelTrue, "SE") <- X * 0.0

listModels <- list(
  RetrieveBestModel(PCMFitDemoObjects$fitBM), 
  RetrieveBestModel(PCMFitDemoObjects$fitOU),
  RetrieveBestModel(PCMFitDemoObjects$fitMGPMTrueTypeMapping), 
  RetrieveBestModel(PCMFitDemoObjects$fitMGPMTrueTypeMappingCheat), 
  modelTrue,
  bestFit$inferredModel)

dtSummary <- data.table(
  model = c(
    "Global BM", 
    "Global OU", 
    "True MGPM, known shifts, unknown parameters", 
    "True MGPM, known shifts, starting from known true parameters (cheating)", 
    "True MGPM, fixed true shifts and true parameters",
    "Inferred MGPM, unknown shifts and parameters"),
  p = sapply(listModels, PCMParamCount),
  logLik = sapply(listModels, logLik), 
  AIC = sapply(listModels, AIC))
knitr::kable(dtSummary)
```

```{r, fig.width=7, fig.height=7, fig.cap="**Example phylogenetic comparative data.** **A**: a tree of 80 tips partitioned in three evolutionary regimes. Each evolutionary regime is denoted by #.T, where # is the regime identifier and T is the evolutionary model type associated with this regime (a model type among A, ..., F). **B**:  bivariate trait values at the tips of the tree. **C** and **D**: Analogical to **A** and **B** according to the MGPM found using the AIC score as a criterion.", echo=FALSE}
library(ggtree)
library(ggplot2)

TipLabelTable <- function(model, ...) {
  tree <- PCMTree(attr(model, "tree"))
  data.table(
    node = sapply(sapply(PCMRegimes(tree), 
                         PCMTreeGetTipsInRegime, 
                         tree = tree), sample, size = 1),
    part.model = 
      paste0(" ", PCMRegimes(tree), ".", 
             LETTERS[PCMMapModelTypesToRegimes(model, tree)], " "), 
    ...)
}

plTree <- PCMTreePlot(attr(modelTrue, "tree"), layout="fan") %<+% 
  TipLabelTable(modelTrue, offset = 5) + 
  geom_tiplab2(size = 2) + 
  geom_tiplab2(aes(label = part.model), offset = 16) + 
  geom_nodelab(size = 2, color = "black") + 
  geom_treescale(
    width = max(PCMTreeNodeTimes(attr(modelTrue, "tree"))), x = 0, 
    linesize = .25, fontsize = 2, offset = 79)

plX <- PCMPlotTraitData2D(
  X[, seq_len(PCMTreeNumTips(attr(modelTrue, "tree")))], 
  attr(modelTrue, "tree"), 
  scaleSizeWithTime = FALSE,
  numTimeFacets = 4, nrowTimeFacets = 2, ncolTimeFacets = 2) +
  geom_text(
    aes(x = x, y = y, label = id, color = regime), 
    size=2, 
    position = position_jitter(.4, .4)) +
  theme_bw() +
  theme(legend.position = "bottom")

plTreeInferred <- 
  PCMTreePlot(attr(bestFit$inferredModel, "tree"), layout="fan") %<+% 
  TipLabelTable(bestFit$inferredModel, offset = 5) +
  geom_tiplab2(size = 2) + 
  geom_tiplab2(aes(label = part.model), offset = 16) + 
  geom_nodelab(size = 2, color = "black") + 
  geom_treescale(
    width = max(PCMTreeNodeTimes(attr(bestFit$inferredModel, "tree"))), x = 0, 
    linesize = .25, fontsize = 2, offset = 79)

plXInferred <- PCMPlotTraitData2D(
  X[, seq_len(PCMTreeNumTips(attr(bestFit$inferredModel, "tree")))], 
  attr(bestFit$inferredModel, "tree"), 
  scaleSizeWithTime = FALSE,
  numTimeFacets = 4, nrowTimeFacets = 2, ncolTimeFacets = 2) +
  geom_text(
    aes(x = x, y = y, label = id, color = regime), 
    size=2, 
    position = position_jitter(.4, .4)) +
  theme_bw() +
  theme(legend.position = "bottom")

cowplot::plot_grid(
  plTree, plX, plTreeInferred, plXInferred, labels = LETTERS[1:4], nrow = 2)
```

# References
